Title

### 2014-01-22 paper/parts/methods.tex
Added labels for equations.
Merge branch 'master' of https://github.com/ddarmon/sfi-dynComm
Conflicts:
	paper/paper.tex

In the following sections, we introduce the problem of community detection, and present the data set used for our analyses. We then describe our methodology for constructing the question-specific networks under study. In particular, we introduce an information theoretic method for activity-based communities, a retweet-mention statistic for interaction-based communities, and a hashtag similarity metric for defining topic-based communities.

@@

%[[Already defined]]In network theory communities are usually defined as group of %nodes more densely connected among each other than with the rest of the network.
As discussed in the introduction, we adopt the standard definition of \emph{community}: a collection of nodes (users) within a network who are more densely connected to each other than with the rest of the network. Structural community detection is a well studied problem and several different methods and algorithms have been proposed; for a complete review of this subject we refer the reader to \cite{fortunato2010community}. In this paper however we focus on a class of networks and communities that is far less studied, in particular we study networks which are both \textit{weighted} and \textit{directed} and communities within those weighted directed networks that can (but need not) \emph{overlap}. When selecting a detection algorithm we propose that all three (weight, direction and overlap) are important for the following reasons: Twitter is a directed network and we wish to maintain this structure. But we are interested in not just structure but function, to capture function we use edge weightings which must be incorporated into the community detection process. Finally, since people can belong to multiple and possibly overlapping social (e.g., college friends, co-workers, family, etc.) and topical (e.g., a user can be interested in both cycling and politics and use the network to discuss the two topics with the two different communities) communities, we are interested in finding \textit{overlapping} communities, rather than partitions of the weighted directed network. 
%For example, users can belong to multiple social groups such as college friends, %co-workers, family, etc. and these groups may but need not overlap. A user can also %belong to multiple topical communities, e.g., a user can be interested in both %cycling and politics and use the network to discuss the two topics with the two %different communities.

This poses an interesting problem because the majority of community detection algorithms developed so far are built to find partitions of a network and very few are aimed at finding overlapping communities \cite{BaumesGKMP05,PalEtAl05,ZhaWanZha07,Gre07,PhysRevE.77.016107,Lancichinetti2009,PhysRevE.80.016105,Kovacs2010}. Among these methods, even fewer deal with directed or weighted networks. For example, the work of~\cite{PalEtAl05} on clique percolation can account for both features, but not at the same time. A recent method proposed by~\cite{LancichinettiPlos}, OSLOM (Order Statistics Local Optimization Method), is one of the first methods  able to deal with all of these features simultaneously. Their method relies on a fitness function that measures the statistical significance of clusters with respect to random fluctuations, and attempts to optimize this fitness function across all clusters. For this paper we use OSLOM as it allowed us to detect \emph{overlapping} communities present in our \emph{weighted} and \emph{directed} network.

@@

Since our goal is to explore the functional communities of this network, we filter the network down to the subset of users which actively interact with each other (e.g., via retweets and mentions). We do this through a metric we call (incoming/outgoing) information events. We define an outgoing information event for a given user $u$ as either a mention made by $u$ of another user in the network, or a retweet of one of $u$'s tweets by another user in the network. The logic for this definition is as follows: if $u$ mentions a user $v$ this can be thought of as $u$ directly sending information to $v$, and if $u$ is retweeted by $v$ then $v$ received information from $u$ and rebroadcast it to their followers. In either case there was information outgoing from $u$ which affected the network in the some way. Analogously, we define the incoming information event for $u$ as either being mentioned by a different user in the network, or as retweeting another user in the network.
%function and information share is important We then filter down this starting %network to include only users that [[what??]]. To do this 
%Since one of the community types we want to explore is based on the explicit interactions ( between users, we filter the dataset in order to take into account only users who make use of these kinds of features in their tweets. [[Seeems out of place]]
%We define an event of outgoing information for a given user $u$ as either a mention made by $u$ of another user in the network, or a retweet by another user of one of $u$'s tweets. 
%[[Switched the readers perspective]]When we mention someone we are in fact sending him some information, and when we are retweeted it means that the person that retweeted us has received some information from us and is sharing it. 
With (incoming/outgoing) information events defined we filtered the network by eliminating all users with less than 9 outgoing \emph{and} incoming information events, i.e., less than one information event per type per week on average. 
We then further restricted our analysis to the strong giant connected component of the network built from the (incoming/outgoing) information filtered set of users. %and whose link represent a user-follower relationship.
In this study the link is directed from the user to the follower because this is the direction in which the information (in the case of transfer entropy) or influence (in the case of mention-retweets) flows. The final network consists of 6,917 nodes and 1,481,131 edges.

@@

For the communities based on transfer entropy, we weight each edge from a user $u$ to a follower $f$ by the estimated transfer entropy of the user $u$ on $f$, 
\begin{align}
	w_{u \to f}^{\text{TE}(k)} = \widehat{\text{TE}}_{X(u) \to X(f)}^{(k)}.
\end{align}

Operationally, we expect users to interact with Twitter on a human time scale, and thus the natural one-second time resolution is too fine. We coarsen each time series by considering non-overlapping time intervals ten minutes in length. For each time interval, we record a 1 if the user has tweeted during that time interval, and a 0 if they have not. Thus, the new coarsened time series now captures whether or not the user has been active on Twitter over any given ten minute time interval in our data set. We then compute the transfer entropy on these coarsened time series taking $k$ to range from 1 to 6, which corresponds to a lag of ten minutes to an hour. The choice of lag must balance a trade-off between additional information and sparsity of samples: as we increase the lag, we account for longer range dependencies, but we also decrease the number of samples availabe to infer a higher dimensional predictive distribution. Ultimately, due to similarities in the underlying communities (see Figure \ref{Fig-compare_coverings}), we chose a lag-4 transfer entropy. All references to activity-based weights, unless otherwise noted, refer to this choice of lag.

@@

Retweets and mentions are two useful features of Twitter networks which can be used to track information flow through the network.
%A way of tracking the flow of information through users is by considering two useful features of this social network: mentions and retweets. 
With mentions users are sending directed information to other users and with retweets users are rebroadcasting information from a user they follow to all of their followers. 
%information from a user has been captured by a follower and shared with his/her own followers. 
This type of information flow defines a community in a much different way than transfer entropy. Instead of defining communities by the loss of uncertainty in one user's tweeting history based on another's, we define an interaction-based communities by weighting the user-follower network with a measure proportional to the number of mentions and/or retweets between users. 
% We can therefore define interaction-based communities by weighting the user-follower network with a measure proportional to the number of mentions and retweets between users. 
For the interaction-based communities we consider three weighting schemes: proportional retweets,
\begin{equation}
w_{u \to f}^{\text{R}}=pR=\frac{\mbox{\# retweets of }u \mbox{ by }f}{\mbox{\# total retweets made by }f},
\end{equation}
proportional mentions,
\begin{equation}
w_{u \to f}^{\text{M}} = pM = \frac{\mbox{\# mentions of }f \mbox{ by }u}{\mbox{\# total mentions of }f},
\end{equation}
and mention-retweet as the arithmetic mean of these two measures,
%we weight each edge from a user $u$ to a follower $f$ through a combination of proportional mentions ($pM$):
% In more details, for each couple of user-follower $u$ and $f$, we use the arithmetic mean of the two following numbers:
%\begin{equation}
%w_{u \to f}^{\text{M}} = pM = \frac{\mbox{\# mentions of }f \mbox{ by }u}{\mbox{\# %total mentions of }f}
%\end{equation}
\begin{align}
	w_{u \to f}^{\text{MR}} = \frac{(pM+pR)}{2}.
\end{align}

@@

For the topical communities we weight each directed edge from a user $u$ to a follower $f$ with the cosine similarity of their respective vectors $\vec{h}(u)$ and $\vec{h}(f)$:
\begin{align}
	w_{u \to f}^{\text{HT}} = \frac{\vec{h}(u) \cdot \vec{h}(f)}{||\vec{h}(u)|| \ \ ||\vec{h}(f)||}.
\end{align}
This weight captures the similarity between users in terms of the topics discussed in their tweets.

>>>>

In the following sections, we introduce the problem of community detection, and present the data set used for our analyses. We then describe our methodology for constructing the question-specific networks. In particular, we introduce an information theoretic method for activity-based communities, a retweet-mention statistic for interaction-based communities, and a hashtag similarity metric for defining topic-based communities.

@@

%[[Already defined]]In network theory communities are usually defined as group of %nodes more densely connected among each other than with the rest of the network.
As discussed in the introduction, we adopt the standard definition of \emph{community}: a collection of nodes (users) within a network who are more densely connected to each other than with the rest of the network. Structural community detection is a well studied problem and several different methods and algorithms have been proposed; for a complete review of this subject we refer the reader to \cite{fortunato2010community}. In this paper however we focus on a class of networks and communities that is far less studied, in particular we study networks which are both \textit{weighted} and \textit{directed} and communities within those weighted directed networks that can (but need not) \emph{overlap}. When selecting a detection algorithm we propose that all three (weight, direction, and overlap) are important for the following reasons. First, communication on Twitter occurs in a directed manner, with users broadcasting information to their followers. An undirected representation of the network would ignore this fact, and could lead to communities composed of users who do not actually share information. Second, we are interested in not just the structure of links but also in their function, and to capture this we use edge weightings which must be incorporated into the community detection process. Finally, since people can belong to multiple and possibly overlapping social (e.g., college friends, co-workers, family, etc.) and topical (e.g., a user can be interested in both cycling and politics and use the network to discuss the two topics with the two different communities) communities, we are interested in finding \textit{overlapping} communities, rather than partitions of the weighted directed network. 
%For example, users can belong to multiple social groups such as college friends, %co-workers, family, etc. and these groups may but need not overlap. A user can also %belong to multiple topical communities, e.g., a user can be interested in both %cycling and politics and use the network to discuss the two topics with the two %different communities.

This last criterion in particular poses a problem because the majority of community detection algorithms developed so far are built to find partitions of a network and few are aimed at finding overlapping communities \cite{BaumesGKMP05,PalEtAl05,ZhaWanZha07,Gre07,PhysRevE.77.016107,Lancichinetti2009,PhysRevE.80.016105,Kovacs2010}. Among these methods, even fewer deal with directed or weighted networks. For example, the work of~\cite{PalEtAl05} on clique percolation can account for both features, but not at the same time. A recent method proposed by~\cite{LancichinettiPlos}, OSLOM (Order Statistics Local Optimization Method), is one of the first methods  able to deal with all of these features simultaneously. Their method relies on a fitness function that measures the statistical significance of clusters with respect to random fluctuations, and attempts to optimize this fitness function across all clusters. For this paper we use OSLOM since it allowed us to detect \emph{overlapping} communities present in our \emph{weighted} and \emph{directed} network.

@@

Since our goal is to explore the functional communities of this network, we filter the network down to the subset of users which actively interact with each other (e.g., via retweets and mentions). We do this by measuring what we call (incoming/outgoing) information events. We define an outgoing information event for a given user $u$ as either a mention made by $u$ of another user in the network, or a retweet of one of $u$'s tweets by another user in the network. The logic for this definition is as follows: if $u$ mentions a user $v$ this can be thought of as $u$ directly sending information to $v$, and if $u$ is retweeted by $v$ then $v$ received information from $u$ and rebroadcast it to their followers. In either case there was information outgoing from $u$ which affected the network in the some way. Analogously, we define the incoming information event for $u$ as either being mentioned by a different user in the network, or as retweeting another user in the network.
%function and information share is important We then filter down this starting %network to include only users that [[what??]]. To do this 
%Since one of the community types we want to explore is based on the explicit interactions ( between users, we filter the dataset in order to take into account only users who make use of these kinds of features in their tweets. [[Seeems out of place]]
%We define an event of outgoing information for a given user $u$ as either a mention made by $u$ of another user in the network, or a retweet by another user of one of $u$'s tweets. 
%[[Switched the readers perspective]]When we mention someone we are in fact sending him some information, and when we are retweeted it means that the person that retweeted us has received some information from us and is sharing it. 
With (incoming/outgoing) information events defined we filtered the network by eliminating all users with less than 9 outgoing \emph{and} incoming information events, i.e., less than one information event per type per week on average. 
We then further restricted our analysis to the strong giant connected component of the network built from the (incoming/outgoing) information filtered set of users. %and whose link represent a user-follower relationship.
In this study the link is directed from the user to the follower because this is the direction in which the information (in the case of transfer entropy) or influence (in the case of mention-retweets) flows. Thus, for a pair of users $u$ and $v$, an edge $a_{v \to u}$ in the structural network has weight 1 if user $u$ follows $v$, and 0 otherwise. The final network consists of 6,917 nodes and 1,481,131 edges.

@@

For the communities based on transfer entropy, we weight each edge from a user $u$ to a follower $f$ by the estimated transfer entropy of the user $u$ on $f$, 
\begin{align}
	w_{u \to f}^{\text{TE}(k)} = \widehat{\text{TE}}_{X(u) \to X(f)}^{(k)}. \label{Eqn-EW-activity}
\end{align}

Operationally, we expect users to interact with Twitter on a human time scale, and thus the natural one-second time resolution is too fine since most humans do not write tweets at the time scale of seconds. We coarsen each time series by considering non-overlapping time intervals ten minutes in length. For each time interval, we record a 1 if the user has tweeted during that time interval, and a 0 if they have not. Thus, the new coarsened time series now captures whether or not the user has been active on Twitter over any given ten minute time interval in our data set. We then compute the transfer entropy on these coarsened time series taking $k$ to range from 1 to 6, which corresponds to a lag of ten minutes to an hour. The choice of lag must balance a trade-off between additional information and sparsity of samples: as we increase the lag, we account for longer range dependencies, but we also decrease the number of samples availabe to infer a higher dimensional predictive distribution. Ultimately, due to similarities in the underlying communities we chose a lag-4 transfer entropy. All references to activity-based weights, unless otherwise noted, refer to this choice of lag.

@@

Retweets and mentions are two useful features of Twitter networks which can be used to track information flow through the network.
%A way of tracking the flow of information through users is by considering two useful features of this social network: mentions and retweets. 
With mentions users are sending directed information to other users and with retweets users are rebroadcasting information from a user they follow to all of their followers. 
%information from a user has been captured by a follower and shared with his/her own followers. 
This type of information flow defines a community in a much different way than transfer entropy. Instead of defining communities by the loss of uncertainty in one user's tweeting history based on another's, we define interaction-based communities by weighting the user-follower network with a measure proportional to the number of mentions and/or retweets between users. 
% We can therefore define interaction-based communities by weighting the user-follower network with a measure proportional to the number of mentions and retweets between users. 
For the interaction-based communities we consider three weighting schemes: proportional retweets,
\begin{equation}
w_{u \to f}^{\text{R}}=pR=\frac{\mbox{\# retweets of }u \mbox{ by }f}{\mbox{\# total retweets made by }f},
\end{equation}
proportional mentions,
\begin{equation}
w_{u \to f}^{\text{M}} = pM = \frac{\mbox{\# mentions of }f \mbox{ by }u}{\mbox{\# total mentions of }f},
\end{equation}
and mention-retweet as the arithmetic mean of these two measures,
%we weight each edge from a user $u$ to a follower $f$ through a combination of proportional mentions ($pM$):
% In more details, for each couple of user-follower $u$ and $f$, we use the arithmetic mean of the two following numbers:
%\begin{equation}
%w_{u \to f}^{\text{M}} = pM = \frac{\mbox{\# mentions of }f \mbox{ by }u}{\mbox{\# %total mentions of }f}
%\end{equation}
\begin{align}
	w_{u \to f}^{\text{MR}} = \frac{(pM+pR)}{2}. \label{Eqn-EW-interaction}
\end{align}

@@

For the topical communities we weight each directed edge from a user $u$ to a follower $f$ with the cosine similarity of their respective vectors $\vec{h}(u)$ and $\vec{h}(f)$:
\begin{align}
	w_{u \to f}^{\text{HT}} = \frac{\vec{h}(u) \cdot \vec{h}(f)}{||\vec{h}(u)|| \ \ ||\vec{h}(f)||}. \label{Eqn-EW-topic}
\end{align}
This weight captures the similarity between users in terms of the topics discussed in their tweets.

### 2014-01-22 paper/parts/results.tex
Merge branch 'master' of https://github.com/ddarmon/sfi-dynComm
Conflicts:
	paper/paper.tex

The number of singletons by community type is also shown in Table~\ref{Table-comm_count}. We see that the topic- and interaction-based communities have the most singletons, with the activity-based community dominating this measure. This result for the activity-based community is partially an artifact of a property of the retweet/mention weighting: 717 of the users were disconnected from the network by how the weights were defined, resulting in `orphan' nodes which we have included in the collection of singletons for all of our analyses. However, even after accounting for this artifact, the interaction-based network still has the most non-orphan singletons. This seems to indicate that a large fraction of the 6917 (nearly 25\%) do not interact with each other in a concerted way that would mark them as a community under our interaction-based definition.

@@

Thus, we see that although the activity-based, interaction-based, and topic-based communities relied on the structural network, their community structure differs \emph{the most} from the community structure of the follower network. This agrees with the results from the previous section, and reinforces that the follower network is a necessary but not sufficient part of detecting communities in online social networks.

@@

We next explore how edge weights, and thus different forms of information flow, differ between community types. For a fixed community type, edges for a particular community may be partitioned into three sets: those from a user in the community to another user in the community (internal-to-internal), those from a user in the community to a user outside of the community (internal-to-external), and those from a user outside the community to a user inside the community (external-to-internal). See Figure~\ref{Fig-edge_types} for a schematic of this edge partitioning. For a meaningful community, we expect the distribution of weights within the community (internal-to-internal weights) to be different from the distribution of weights without the community (internal-to-external and external-to-internal).

@@

\begin{table}
	\caption{The median value across the 100 largest communities for the ratio of the median internal-to-internal weight to the median external-to-internal / internal-to-external weight for the different community / weight pairings. Community types correspond to S(tructural), A(ctivity-based), T(opic-based), and I(nteraction-based) communities. Weight types correspond to T(ransfer) E(ntropy), M(ention-)R(etweet), and H(ash)T(ag). Note: For mention-retweet weights, zero weight edges were excluded from the computation of the median. We indicate such cases with an asterisk.}
	\centering
	\begin{tabular}{c  c | c  c  c}
		& & \multicolumn{3}{ c }{Weight Type} \\ %\cline{3-5}
		& & TE & MR & HT \\ \cline{1-5}
		\multirow{4}{*}{Community} & S & $0.96/0.94$& $1.7/2.1$*& $9.0/8.0$\\
		\multirow{4}{*}{Type} & A & $1.0/0.96$& $1.5/2.4$*& $24/17$\\
		& I & $0.83/0.86$& $3.2/4.4$& $10/8.5$\\
		& T & $0.9/0.89$& $2.4/2.6$*& $28/26$
	\end{tabular}
	\label{Table-medians}
\end{table}

>>>>

We see in Table~\ref{Table-comm_count} that the topic- and interaction-based communities have the most singletons, with the activity-based community dominating this measure. This result for the activity-based community is partially an artifact of a property of the retweet/mention weighting: 717 of the users were disconnected from the network by how the weights were defined, resulting in `orphan' nodes which we have included in the collection of singletons for all of our analyses. However, even after accounting for this artifact, the interaction-based network still has the most non-orphan singletons. This seems to indicate that a large fraction of the 6917 (nearly 25\%) do not interact with each other in a concerted way that would mark them as a community under our interaction-based definition. This agrees with a result previously reported in~\cite{romero2011influence} about how most users passively interact with incoming information on Twitter.

@@

Thus, we see that although the activity-based, interaction-based, and topic-based communities relied on the structural network, their community structure differs \emph{the most} from the community structure of the follower network. This agrees with the results from the previous section, and reinforces that the follower network is a necessary but not sufficient part of detecting communities characterized by properties beyond follower-followee relationships.

@@

We next explore how the edge weights defined by equations (\ref{Eqn-EW-activity}), (\ref{Eqn-EW-interaction}), and (\ref{Eqn-EW-topic}), and thus different forms of information flow, differ between community types. For a fixed community type, edges for a particular community may be partitioned into three sets: those from a user in the community to another user in the community (internal-to-internal), those from a user in the community to a user outside of the community (internal-to-external), and those from a user outside the community to a user inside the community (external-to-internal). See Figure~\ref{Fig-edge_types} for a schematic of this edge partitioning. For a meaningful community, we expect the distribution of weights within the community (internal-to-internal weights) to be different from the distribution of weights without the community (internal-to-external and external-to-internal).

@@

\begin{table}
	\caption{The median value across the 100 largest communities for the ratio of the median internal-to-internal weight to the median external-to-internal / internal-to-external weight for the different community / weight pairings. For each entry $a/b$ in the table, $a$ corresponds to median ratio value for edges external-to-internal, and $b$ corresponds to the median ratio value for edges internal-to-external . Community types correspond to S(tructural), A(ctivity-based), T(opic-based), and I(nteraction-based) communities. Weight types correspond to T(ransfer) E(ntropy), M(ention-)R(etweet), and H(ash)T(ag). Note: For mention-retweet weights, zero weight edges were excluded from the computation of the median. We indicate such cases with an asterisk.}
	\centering
	\begin{tabular}{c  c | c  c  c}
		& & \multicolumn{3}{ c }{Weight Type} \\ %\cline{3-5}
		& & TE & MR & HT \\ \cline{1-5}
		\multirow{4}{*}{Community} & S & $0.96/0.94$& $1.7/2.1$*& $9.0/8.0$\\
		\multirow{4}{*}{Type} & A & $1.0/0.96$& $1.5/2.4$*& $24/17$\\
		& I & $0.83/0.86$& $3.2/4.4$& $10/8.5$\\
		& T & $0.9/0.89$& $2.4/2.6$*& $28/26$
	\end{tabular}
	\label{Table-medians}
\end{table}
